---
title: "Data Wrangling 2 - Titanic data"
author: "Ivy Chiang"
date: "13 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

##### Workspace setup and file loading

```{r results='hide'}
setwd("~/Foundations/3-DataWrangling/3-Ex2")
ls()
rm(list=ls())
library(tidyr)
library(dplyr)

df <-
read.csv("titanic_original.csv", stringsAsFactors = FALSE)
df <- tbl_df(df)
```

##### 1: Port of embarkation

The embarked column has some missing values, which are known to correspond to passengers who actually embarked at Southampton. Find the missing values and replace them with S. (Caution: Sometimes a missing value might be read into R as a blank or empty string.)

```{r results='hide}
#final attempt
embarked<- df$embarked

#check missing data
original <- length(df$embarked)
any(is.na(embarked))
any(embarked=="")

#Fill missing Data
embarked_c <-sapply(embarked, function(x){ if (x ==""){x <- "S"}else {str_trim(x)}})

#check Fill operation
any(embarked_c=="")
original == length(embarked_c)

#Apply cleaned data to dataset
ans1<- mutate(df, embarked = embarked_c)

#check
any(ans1$embarked=="")
ans1

```

```{r results='hide'}
#second attempt
library(stringr)

embarked<- df$embarked

#check missing data
original <- length(df$embarked)
any(is.na(embarked))
any(embarked=="")
n <- which(embarked=="")

#Fill values
sub("", "S", embarked[n])

#check Fill operation
original == length(embarked)

#first attempt
#Find null values and action value replace
ans1 <- df %>% select(embarked == "") %>% mutate(embarked = "S")

#Trim whitespace and find if more null values are found
library(stringr)
str_trim(ans1$embarked)
ans1 %>% filter(embarked == "")
#no more null values


#gsub("[[:space:]]", "", df$embarked)
#gsub("^\\s+|\\s+$", "", df$embarked)

```

##### 2: Age

You’ll notice that a lot of the values in the Age column are missing. While there are many ways to fill these missing values, using the mean or median of the rest of the values is quite common in such cases.
1. Calculate the mean of the Age column and use that value to populate the missing values
2. Think about other ways you could have populated the missing values in the age column. Why would you pick any of those over the mean (or not)?

*Using mean values will not skew the dataset whilst avoid any mathematical calculations errors 

```{r results='hide'}
#setup variables
mean_age <- mean(ans1$age)
original_a <- length(ans1$age)

#Fill missing values with mean
age <- ans1$age
age_c <-sapply(age, function(x){ if (x =="" | is.na(x)){x <- mean_age}})

#Join cleaned data to dataset
ans2 <- mutate(ans1, age = age_c)

#Check
length(ans2$age) == original_a
ans2

#Past attempt
#ans1 %>% filter(age == "") %>% mutate(age = mean(ans1$age))


```

##### 3: Lifeboat

You’re interested in looking at the distribution of passengers in different lifeboats, but as we know, many passengers did not make it to a boat :-( This means that there are a lot of missing values in the boat column. Fill these empty slots with a dummy value e.g. the string 'None' or 'NA'

```{r results='hide'}
#setup variables
dummy_value <- "None"
boat <- ans2$boat

#Missing values check
sum(any(is.na(boat)))
sum(any(boat == ""))

#Fill missing values with mean
boat_c <-sapply(boat, function(x){ if (x =="" | is.na(x)){x <- dummy_value}})

#Join cleaned data to dataset
ans3 <- mutate(ans2, boat = boat_c)

#Check cells maintained and number of missing values
length(ans3$boat) == length(df$boat)
sum(any(ans3$boat=="" | is.na(ans3$boat)))
ans3

```

##### 4: Cabin

You notice that many passengers don’t have a cabin number associated with them.

* Does it make sense to fill missing cabin numbers with a value?
*Yes, but the filling will be to indicate that no number is recorded so data is more understandable to future readers
* What does a missing value here mean?
* the missing value means the data was not captured, these could be stowaways or workers.

You have a hunch that the fact that the cabin number is missing might be a useful indicator of survival. Create a new column has_cabin_number which has 1 if there is a cabin number, and 0 otherwise.


```{r results='hide'}
#checking hypothesis of no cabin value meaning survival
df %>% 
  select(survived, cabin) %>% 
  filter(cabin == "")

#Create list
new <- unlist(lapply(ans3$cabin, function(x) {if (x =="" | is.na(x)){return (0)}else {return(1)}}))

ans4<- mutate(ans3, has_cabin_number = new )

head(ans4$has_cabin_number, n=10)

```

##### 6: Submit the project on Github

Include your code, the original data as a CSV file titanic_original.csv, and the cleaned up data as a CSV file called titanic_clean.csv.

```{r results='hide'}
ans6<- as.matrix(ans4)
write.table(ans6,"titanic_clean.csv", sep=",",row.names = FALSE, qmethod = "double")


```
